% -----------------------------------------------------------------------------
% File: sections/03_alzheimer_diagnosis.tex
% Description: Section 3 - Alzheimer Diagnosis Based on VGG16 Fusion
% -----------------------------------------------------------------------------

\section{Paper 2: Alzheimer Diagnosis Based on VGG16 Fusion}
\label{sec:alzheimer}

\paperCard{The New Method for Detection of Alzheimer's Disease}{Bartosz Brejna, Kacper Szmerga\l{}a, Adrianna Kozierkiewicz}{ICCCI 2025, LNAI 16138, pp. 346--360}

\subsection{Background and Research Motivation}
Alzheimer's Disease (AD) is a progressive neurodegenerative disorder leading to severe cognitive decline, memory loss, and brain tissue atrophy. As the global population aged 65+ is projected to double by 2050, the demand for early and accurate AD diagnosis has become a pressing public health priority. Early detection is critical for administering disease-modifying therapies that can slow disease progression.

Magnetic Resonance Imaging (MRI) is a primary non-invasive diagnostic modality used to detect structural brain changes, particularly hippocampal atrophy, which is a hallmark of Alzheimer's pathology. However, manual interpretation of MRI scans is time-consuming, requires specialized expertise, and is prone to inter-observer variability, especially given the projected shortage of approximately 42,000 radiologists by 2033.

A critical methodological flaw in many published high-accuracy deep learning studies is \highlight{data leakage} during dataset splitting. If a dataset is split at the 2D slice level, slices from the same patient (subject) can appear in both the training and testing sets. Because consecutive MRI slices of a patient are highly correlated, the model memorizes patient-specific anatomical features rather than learning general pathological indicators, leading to overoptimistic performance that fails in clinical practice. To prevent this, a strict \highlight{subject-wise split} must be enforced, ensuring that all slices from one patient belong exclusively to either the training or testing set.

\subsection{Paper Objectives}
The paper proposes a robust diagnostic pipeline for Alzheimer's Disease detection that addresses the data leakage problem. The specific objectives include:
\begin{itemize}
    \item Implementing a standardized MRI preprocessing pipeline using FSL tools for hippocampal ROI extraction.
    \item Applying progressive transfer learning with VGG16 to prevent catastrophic forgetting on small medical datasets.
    \item Proposing a dual-model disjunction (logical OR) fusion strategy to maximize diagnostic sensitivity and reduce missed diagnoses.
    \item Enforcing subject-wise data splitting to ensure clinically valid performance evaluation.
\end{itemize}

\subsection{Proposed Method}

\subsubsection{Standardized MRI Preprocessing Pipeline}
The raw 3D T1-weighted MRI volumes from the ADNI dataset undergo a standardized six-step preprocessing pipeline to extract the hippocampus Region of Interest (ROI):
\begin{enumerate}
    \item \textbf{FOV Cropping:} Restricting the field-of-view to exclude irrelevant neck tissue.
    \item \textbf{BET (Brain Extraction Tool):} Skull-stripping to isolate brain tissue from surrounding skull and scalp.
    \item \textbf{FLIRT (Linear Registration):} Aligning the brain volume to the MNI152 standard template using a 12-parameter affine transformation.
    \item \textbf{FNIRT (Non-Linear Registration):} Refining the alignment locally using non-linear warping to account for individual anatomical variations.
    \item \textbf{ROI Extraction:} Isolating the hippocampus region to yield $224 \times 224 \times 3$ 2D slices suitable for VGG16 input.
\end{enumerate}

\begin{figure}[h]
    \centering
    \resizebox{0.95\linewidth}{!}{%
    \begin{tikzpicture}[
        node distance=0.5cm and 0.3cm,
        font=\sffamily\tiny,
        stepbox/.style={rectangle, draw=techBlue!80, fill=white, thick, rounded corners=2pt, text width=1.6cm, align=center, inner sep=2pt, drop shadow={opacity=0.3}},
        arrow/.style={->, >=stealth, thick, techBlue, rounded corners=2pt}
    ]
        \node[stepbox] (step1) {\textbf{1. Raw Volume}\\(ADNI Dataset)};
        \node[stepbox, right=of step1] (step2) {\textbf{2. FOV Crop}\\(Cropping)};
        \node[stepbox, right=of step2] (step3) {\textbf{3. BET}\\(Skull Stripping)};
        \node[stepbox, below=0.7cm of step3] (step4) {\textbf{4. FLIRT}\\(Linear Reg.)};
        \node[stepbox, left=of step4] (step5) {\textbf{5. FNIRT}\\(Non-Lin Reg.)};
        \node[stepbox, left=of step5] (step6) {\textbf{6. ROI Extraction}\\(Hippocampus)};
        \draw[arrow] (step1) -- (step2);
        \draw[arrow] (step2) -- (step3);
        \draw[arrow] (step3.south) -- ++(0,-0.3) -| (step4.north);
        \draw[arrow] (step4) -- (step5);
        \draw[arrow] (step5) -- (step6);
    \end{tikzpicture}%
    }
    \caption{Six-step standardized MRI preprocessing pipeline utilizing FSL tools to extract the hippocampal ROI.}
    \label{fig:mri_pipeline}
\end{figure}

\subsubsection{Data Augmentation Strategy}
To address the risk of overfitting on small medical datasets, the authors apply online data augmentation during training. Transformations include horizontal flipping ($p=0.5$), random rotation ($\pm 10\%$ radians), translation ($\pm 10\%$ shift), zoom ($\pm 10\%$ scale), and contrast adjustment ($\pm 10\%$ intensity). These are applied on-the-fly to force the model to learn features invariant to position, orientation, and intensity variations.

\subsubsection{Progressive Transfer Learning with VGG16}
To prevent catastrophic forgetting and stabilize training on small datasets, a three-phase fine-tuning schedule is implemented:
\begin{itemize}
    \item \textbf{Phase 1 (Feature Extraction):} The entire VGG16 backbone is frozen; only the custom classifier head is trained ($\eta = 10^{-4}$, 10 epochs).
    \item \textbf{Phase 2 (Shallow Fine-tuning):} The last 8 layers of the backbone are unfrozen and trained with a reduced learning rate ($\eta = 10^{-5}$, 10 epochs).
    \item \textbf{Phase 3 (Deep Fine-tuning):} The last 10 layers are unfrozen for full domain adaptation ($\eta = 10^{-5}$, 20 epochs).
\end{itemize}

\subsubsection{Dual-Model Disjunction Fusion}
To mitigate false negatives in clinical diagnostics, the authors propose a logical disjunction (OR) fusion strategy. Two models are trained: a Base VGG16 (frozen backbone) and a Tuned VGG16 (fine-tuned backbone). Let $L_1$ be the binary prediction of the Base model and $L_2$ be the prediction of the Tuned model:
\begin{equation}
    L_{final} = L_1 \lor L_2
\end{equation}
If \textit{either} model flags the MRI as positive for AD, the final diagnosis is positive. This strategy prioritizes sensitivity over precision, which is critical in clinical settings where missing a positive case (false negative) can delay treatment.

\begin{figure}[h]
    \centering
    \resizebox{0.9\linewidth}{!}{%
    \begin{tikzpicture}[
        node distance=0.5cm and 0.8cm,
        font=\sffamily\small,
        input/.style={rectangle, draw=black, thick, fill=gray!10, minimum width=1.2cm, minimum height=1.2cm, rounded corners=2pt, align=center},
        model/.style={rectangle, draw=techBlue, thick, fill=white, minimum width=2.2cm, minimum height=0.7cm, rounded corners=3pt, drop shadow, align=center},
        gate/.style={circle, draw=techBlue, ultra thick, fill=white, minimum size=1.0cm, inner sep=0pt, align=center},
        outcome/.style={rectangle, draw=black, thick, minimum width=1.8cm, minimum height=0.7cm, rounded corners=3pt, font=\bfseries, align=center},
        arrow/.style={->, >=stealth, ultra thick, techBlue, rounded corners=5pt}
    ]
        \node[input] (mri) {Subject\\MRI};
        \node[model, above right=0.0cm and 1.2cm of mri] (m1) {Base VGG16 ($L_1$)};
        \node[model, below right=0.0cm and 1.2cm of mri] (m2) {Tuned VGG16 ($L_2$)};
        \coordinate (midpoint) at ($(m1.east)!0.5!(m2.east)$);
        \node[gate, right=1.5cm of midpoint] (or) {\Huge $\lor$};
        \node[outcome, above right=0.1cm and 1.2cm of or, draw=alertRed, fill=red!10, text=alertRed] (ad) {AD (1)};
        \node[outcome, below right=0.1cm and 1.2cm of or, draw=codeGreen, fill=green!10, text=codeGreen] (cn) {CN (0)};
        \draw[arrow] (mri.east) -- ++(0.4,0) |- (m1.west);
        \draw[arrow] (mri.east) -- ++(0.4,0) |- (m2.west);
        \draw[arrow] (m1.east) -- node[midway, above, font=\scriptsize] {$L_1$} ++(0.6,0) |- (or.west);
        \draw[arrow] (m2.east) -- node[midway, above, font=\scriptsize] {$L_2$} ++(0.6,0) |- (or.west);
        \draw[arrow] (or.east) -- ++(0.6,0) |- node[pos=0.7, right, font=\scriptsize, text=alertRed] {True} (ad.west);
        \draw[arrow, dashed, gray] (or.east) -- ++(0.6,0) |- node[pos=0.7, right, font=\scriptsize, text=codeGreen] {False} (cn.west);
    \end{tikzpicture}%
    }
    \caption{Dual-model logical disjunction (OR) fusion pipeline. If either model predicts AD, the final diagnosis is positive.}
    \label{fig:fusion_logic}
\end{figure}

\subsubsection{Subject-Level Inference}
For each patient, the model aggregates predictions from two selected hippocampal slices. The subject-level prediction probability is computed as:
\begin{equation}
    P_{subject} = \frac{p_1 + p_2}{2} \quad \Rightarrow \quad \hat{y} = \begin{cases} \text{AD} & \text{if } P_{subject} > 0.5 \\ \text{CN} & \text{otherwise} \end{cases}
\end{equation}
where $p_1$ and $p_2$ are the prediction probabilities for the two slices.

\subsection{Experiments and Results}

\subsubsection{ADNI Dataset}
The Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset provides T1-weighted MRI scans for subjects classified as Cognitively Normal (CN) or Alzheimer's Disease (AD). The dataset is split at the subject level to prevent data leakage, ensuring that all slices from one patient appear exclusively in either the training or testing partition.

\subsubsection{Experimental Results}
Table~\ref{tab:alzheimer_results} presents the performance metrics for the three model strategies. The disjunction fusion achieves the highest accuracy of \highlight{89.87\%} and boosts diagnostic sensitivity to \highlight{87.50\%} (+12.5\% over the base model), recovering 5 patients who were previously misdiagnosed as healthy.

\begin{table}[h]
    \centering
    \caption{Performance results of VGG16 models and fusion strategy.}
    \label{tab:alzheimer_results}
    \resizebox{\linewidth}{!}{%
    \begin{tabular}{lcccc}
        \toprule
        \textbf{Model Strategy} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Sensitivity} & \textbf{F1-Score} \\
        \midrule
        Base VGG16 ($L_1$) & 83.54\% & 90.91\% & 75.00\% & 82.19\% \\
        Tuned VGG16 ($L_2$) & 87.34\% & \textbf{94.12\%} & 80.00\% & 86.49\% \\
        \rowcolor{techBlue!10}
        \textbf{Fused VGG16 ($L_1 \lor L_2$)} & \textbf{89.87\%} & 92.11\% & \textbf{87.50\%} & \textbf{89.74\%} \\
        \bottomrule
    \end{tabular}
    }
\end{table}

Detailed analysis of the results:
\begin{itemize}
    \item \textbf{Base VGG16 ($L_1$):} Achieves 83.54\% accuracy with high precision (90.91\%) but limited sensitivity (75.00\%), indicating that 10 AD patients are missed (false negatives).
    \item \textbf{Tuned VGG16 ($L_2$):} Progressive fine-tuning improves accuracy to 87.34\% and achieves the highest precision (94.12\%), but sensitivity remains at 80.00\%.
    \item \textbf{Fused VGG16 ($L_1 \lor L_2$):} The logical OR fusion recovers 5 additional true positive patients (TP: $30 \to 35$), boosting sensitivity from 75.00\% to 87.50\% at the cost of only a minor drop in precision (from 94.12\% to 92.11\%).
    \item The fusion strategy demonstrates that combining complementary models via logical disjunction effectively reduces missed diagnoses, which is the most critical clinical requirement.
\end{itemize}

\subsection{Conclusion and Contributions}
The paper proposes a robust pipeline combining standardized MRI preprocessing, progressive transfer learning, and dual-model fusion for Alzheimer's Disease detection.

Key contributions of the paper:
\begin{itemize}
    \item Implements a standardized six-step MRI preprocessing pipeline using FSL tools, ensuring reproducible and clinically valid input data.
    \item Demonstrates progressive transfer learning as an effective strategy for adapting pre-trained models to small medical imaging datasets without catastrophic forgetting.
    \item Proposes the dual-model disjunction fusion ($L_1 \lor L_2$) that achieves 89.87\% accuracy and 87.50\% sensitivity, recovering 5 missed AD patients compared to the baseline.
    \item Enforces subject-wise data splitting to eliminate data leakage, a critical flaw in many prior studies.
    \item Provides a practical framework that balances high sensitivity (critical for clinical safety) with acceptable precision levels.
\end{itemize}
